Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-26 optim.qmd
small typos and abstract

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  Projection pursuit (PP) is a dimension reduction technique that identifies low-dimensional projections in high-dimensional data by optimising a criteria function known as the PP index. The optimisation in PP can be non-smooth and requires identifying optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recent swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effect on JSO success rate. The JSO algorithm has been implemented in the `tourr` package, while calculations for smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - Jellyfish Search Optimiser (JSO)
  - Optimisation
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

This study explores the potential for a Jellyfish Search Optimiser 
(JSO) to be integrated with projection 
pursuit guided tour. JSO [@chou_novel_2021] -[@rajwar_exhaustive_2023], 
inspired by the search behaviour of jellyfish 
in the ocean, is a swarm-based metaheuristic designed to solve global
optimisation problems. Compared to traditional methods, JSO 
demonstrated stronger search ability and faster convergence, and 
requires fewer tuning parameters. These
practical advantages make JSO as a promising candidate for enhancing
PP optimisation. As such, the study aims to explore JSO's potential
in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.

@@

The rest of this paper is structured as follows. @sec-background
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature. @sec-theory details JSO and
introduces metrics that measure different properties of PP indexes.
@sec-vis outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's perforamnce.
@sec-app presents two sets of simulation
experiments, comparing JSO with another optimiser, search better, and
studies the impact of different PP index properties on the optimisation
performance. @sec-conclusion summarises the work and provides
suggestions for future directions.

# Projection pursuit, index functions and optimisation \[Di and Sherry\] {#sec-background}

>>>>

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimizes a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with small a "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated with JSO hyper-parameters to determine their effect on JSO success rate. The JSO algorithm has been implemented in the `tourr` package, while functions to calculate  smoothness and squintability of indexes are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - Jellyfish Search Optimiser (JSO)
  - Optimisation
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

This study explores the potential for a Jellyfish Search Optimiser (JSO) to be integrated with the projection pursuit guided tour. JSO [@chou_novel_2021] -[@rajwar_exhaustive_2023], inspired by the search behaviour of jellyfish in the ocean, is a swarm-based metaheuristic designed to solve global optimisation problems. Compared to traditional methods, JSO demonstrated stronger search ability and faster convergence, and requires fewer tuning parameters. These practical advantages make JSO as a promising candidate for enhancing PP optimisation. As such, the study aims to explore JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.

@@

The rest of this paper is structured as follows. @sec-background introduces the background of PP guided tour and reviews existing optimisers and index functions in the literature. @sec-theory details JSO and introduces metrics that measure different properties of PP indexes. @sec-vis outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's performance. @sec-app presents two sets of simulation experiments, comparing JSO with another optimiser, search better, and studies the impact of different PP index properties on the optimisation
performance. @sec-conclusion summarises the work and provides suggestions for future directions.

# Projection pursuit, tours, index functions and optimisation \[Di and Sherry\] {#sec-background}

### 2024-06-26 optim.tex
small typos and abstract

\begin{abstract}
Projection pursuit (PP) is a dimension reduction technique that
identifies low-dimensional projections in high-dimensional data by
optimising a criteria function known as the PP index. The optimisation
in PP can be non-smooth and requires identifying optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recent
swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising
PP indexes. The performance of JSO is evaluated across various
hyper-parameter settings and compared with existing optimisers.
Additionally, this work proposes novel methods to quantify two
properties of the PP index -- smoothness and squintability -- that
capture the complexities inherent in PP optimisation problems. These two
metrics are evaluated with JSO hyper-parameters to determine their
effect on JSO success rate. The JSO algorithm has been implemented in
the \texttt{tourr} package, while calculations for smoothness and
squintability are available in the \texttt{ferrn} package.
\end{abstract}

@@

This study explores the potential for a Jellyfish Search Optimiser (JSO)
to be integrated with projection pursuit guided tour. JSO
\citep{chou_novel_2021} -\citep{rajwar_exhaustive_2023}, inspired by the
search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO as a promising candidate for enhancing PP
optimisation. As such, the study aims to explore JSO's potential in
enhancing projection pursuit and examines the characteristics and
advantages of such implementation.

@@

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the background of PP guided tour
and reviews existing optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-vis} outlines the
data structures, numerical summaries, and visualisations used in the
simulation studies conducted in Section~\ref{sec-app} to assess JSO's
perforamnce. Section~\ref{sec-app} presents two sets of simulation
experiments, comparing JSO with another optimiser, search better, and
studies the impact of different PP index properties on the optimisation
performance. Section~\ref{sec-conclusion} summarises the work and
provides suggestions for future directions.

\section{Projection pursuit, index functions and optimisation {[}Di and
Sherry{]}}\label{sec-background}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, opacitybacktitle=0.6, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, coltitle=black, titlerule=0mm, breakable, arc=.35mm, toprule=.15mm, left=2mm, bottomrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, colback=white, toptitle=1mm, rightrule=.15mm]

>>>>

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimizes a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with small a
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO is evaluated across
various hyper-parameter settings and compared with existing optimisers.
Additionally, this work proposes novel methods to quantify two
properties of the PP index -- smoothness and squintability -- that
capture the complexities inherent in PP optimisation problems. These two
metrics are evaluated with JSO hyper-parameters to determine their
effect on JSO success rate. The JSO algorithm has been implemented in
the \texttt{tourr} package, while functions to calculate smoothness and
squintability of indexes are available in the \texttt{ferrn} package.
\end{abstract}

@@

This study explores the potential for a Jellyfish Search Optimiser (JSO)
to be integrated with the projection pursuit guided tour. JSO
\citep{chou_novel_2021} -\citep{rajwar_exhaustive_2023}, inspired by the
search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO as a promising candidate for enhancing PP
optimisation. As such, the study aims to explore JSO's potential in
enhancing projection pursuit and examines the characteristics and
advantages of such implementation.

@@

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the background of PP guided tour
and reviews existing optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-vis} outlines the
data structures, numerical summaries, and visualisations used in the
simulation studies conducted in Section~\ref{sec-app} to assess JSO's
performance. Section~\ref{sec-app} presents two sets of simulation
experiments, comparing JSO with another optimiser, search better, and
studies the impact of different PP index properties on the optimisation
performance. Section~\ref{sec-conclusion} summarises the work and
provides suggestions for future directions.

\section{Projection pursuit, tours, index functions and optimisation
{[}Di and Sherry{]}}\label{sec-background}

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, coltitle=black, titlerule=0mm, toptitle=1mm, arc=.35mm, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colback=white]
